```{r, echo=FALSE}
options(warn=-1)
```
The Health and Economic Impacts of Storms in the US since 1950
========================================================

<h2>SYNOPSIS</h2>

The objective of this report is to explore the NOAA Storm Database. 
A particular focus lies on analyzing (1) the impact of storms on health (i.e.,
how many fatalities and injuries are associated with what kind of storm) and
(2) the imapct of storms on the US economy (i.e., how much damage measures in USD
may be associated with what kind of storm). I analyzed the data from 1950 to Nov 2011.

Please note: 
This report is part of my assignment for the "Data Specialization / Reproducible Research" 
lecture offered by John Hopkins University.

The reminder is divided in two main parts. First, in part A I will explain how I processed the data and 
also give you the respective R code. Second, in part B I will show you the results and together with 
three charts that illustrate my findings.


<h2>PART A. DATA PROCESSING</h2>

<h3>1. Install and load packages</h3>

Step 1 installs and loads the required packages. I use data.table for storing and aggregating data and scales, 
xtable and ggplot2 for illustrating my findings and displaying some beautiful charts.

```{r}
##install.packages("data.table")
library(data.table)
##install.packages("scales")
library(scales)
##install.packages("ggplot2")
library(ggplot2)
##install.packages("xtable")
library(xtable)
```

<h3>2. Load Data</h3>

Step 2 reads the data. 

```{r}
data <- read.csv( bzfile("C:/Users/Martin_Admin/Desktop/RepData_PA2/repdata-data-StormData.csv.bz2") )
```

<h3>3. Preprocess data</h3>

Step 3 creates a vector containg all events. This will be helpful for iterating through the data later.

```{r}
eventVector <- as.character( unique(data$EVTYPE) )
```

<h3>4. Task 1: Identification of most harmful events with respect to population health</h3>

Step 4 adresses task 1 of the assignment. It creates a data.table that contains all events and the 
number of facilities and injuries associated with them.

<h4>4.1 Subset data</h4>

Step 4.1 creates a subset. This reduces the processing time later on.

```{r}
dataTask1 <- data.table( data[, c("EVTYPE", "FATALITIES", "INJURIES")] )
```

<h4>4.2 Compute table with health impact</h3>

Step 4.2 creates a new table with one row for each event and then fill the table 
with the number of facilities and injuries. Finally, it orders the table in descending order.

```{r}
healthImpact <- data.table(max1=character(0), max2=integer(0), max3=integer(0), max4=integer(0))
names(healthImpact)[1]<-paste("event")
names(healthImpact)[2]<-paste("eventCount")
names(healthImpact)[3]<-paste("fatalities")
names(healthImpact)[4]<-paste("injuries")

for(i in eventVector){
  
  additionalEventList <- list(i, 0, 0, 0)
  healthImpact <- rbind(healthImpact, additionalEventList)
  
}
## Summ fatalities and injuries for each event
fatalitiesDT <- dataTask1[, sum(FATALITIES), by=EVTYPE]
injuriesDT <- dataTask1[, sum(INJURIES), by=EVTYPE]

## Insert sums of fatalities and injuries in healthImpact data table and order data table
for(i in 1:nrow(healthImpact)){
  
  healthImpact$fatalities[i] <- fatalitiesDT$V1[ fatalitiesDT$EVTYPE==healthImpact$event[i] ]
  healthImpact$injuries[i] <- injuriesDT$V1[ injuriesDT$EVTYPE==healthImpact$event[i] ]
  
}

healthImpact <- healthImpact[order(-fatalities, -injuries)]
```

<h3>5. Task 2: Identification of events with greatest economic consequences</h3>

As mentioned above, step 5 focuses on task 2 of the assignment.

<h4>5.1 Subset data</h4>

Similar to step 4.1, I create a subset called dataTask2 which is specifically tailored to the needs of task 2. 

Furthermore, I also create three tables that will be helpful for the following computations. The table prop 
is a subset of the data that contains data on the economic property damage. Specifically, it contains the event and 
the amound of damage. Interestingly, the amound of damage is stored in two columns. The factor is stored in "PROPDMG" and 
the exponent is stored in "PROPDMGEXP". Accordingly, I create the table crop for the economic damage on crop. 
This table also stores the event, factor, and exponent.

```{r}
dataTask2 <- data.table( data[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")] )

totalDmgDT <- data.table(max1=character(0), max2=numeric(0), max3=numeric(0), max4=numeric(0))
names(totalDmgDT)[1]<-paste("event")
names(totalDmgDT)[2]<-paste("property_dmg")
names(totalDmgDT)[3]<-paste("crop_dmg")
names(totalDmgDT)[4]<-paste("total_dmg")

prop <- dataTask2[, sum(PROPDMG), by=list(EVTYPE, PROPDMGEXP)]
names(prop)[1]<-paste("event")
names(prop)[2]<-paste("exp")
names(prop)[3]<-paste("damage")
prop <- prop[damage>0] ## remove rows where prop damage is zero

crop <- dataTask2[, sum(CROPDMG), by=list(EVTYPE, PROPDMGEXP)]
names(crop)[1]<-paste("event")
names(crop)[2]<-paste("exp")
names(crop)[3]<-paste("damage")
crop <- crop[damage>0] ## remove rows where crop damage is zero
```

<h4>5.2 Clean subsets</h4>

Step 5.2 transforms the exponent columns of the tables prop and crop, because the given data is not 
tidy yet. In particular, I first assign the exponent zero to the characters space, ?, - and +. Second, 
I assign the exponent 2 to H and h, the exponent 3 to k and K, and the exponent 6 to M and m. As the column 
is automatically read as a factor that does not contain the element 9, I cannot assign the exponent 9 in this step. 
Hence, I us a rather quick and dirty approach for this: I simply use an if-statement later on to filer all Bs.


```{r}
for(i in 1:nrow(prop)){
  
  if( as.character(prop$exp[i])=="" | as.character(prop$exp[i])=="?" |as.character(prop$exp[i])=="-" | as.character(prop$exp[i])=="+" ){
    prop$exp[i] <- 0
  }else if( as.character(prop$exp[i])=="H" |as.character(prop$exp[i])=="h" ){ ## hecta
    prop$exp[i] <- 2
  }else if( as.character(prop$exp[i])=="K" |as.character(prop$exp[i])=="k" ){ ## kilo
    prop$exp[i] <- 3
  }else if( as.character(prop$exp[i])=="M" |as.character(prop$exp[i])=="m" ){ ## mil
    prop$exp[i] <- 6
  }else if( as.character(prop$exp[i])=="B" |as.character(prop$exp[i])=="b" ){ ## bil --> not in factor
    prop$exp[i] <- "B"
  }
  
}

for(i in 1:nrow(crop)){
  
  if( as.character(crop$exp[i])=="" | as.character(crop$exp[i])=="?" |as.character(crop$exp[i])=="-" | as.character(crop$exp[i])=="+" ){
    crop$exp[i] <- 0
  }else if( as.character(crop$exp[i])=="H" |as.character(crop$exp[i])=="h" ){
    crop$exp[i] <- 2
  }else if( as.character(crop$exp[i])=="K" |as.character(crop$exp[i])=="k" ){
    crop$exp[i] <- 3
  }else if( as.character(crop$exp[i])=="M" |as.character(crop$exp[i])=="m" ){
    crop$exp[i] <- 6
  }else if( as.character(crop$exp[i])=="B" |as.character(crop$exp[i])=="b" ){
    crop$exp[i] <- "B"
  }
  
}
```

<h4>5.3 Compute damage per event</h4>

Step 5.3 computes fills the table totalDmgDT. This table has one row for each event. 
This row shows the event name, the crop damage, the property damage, and the total economic damage.

```{r}
for(i in 1:length(eventVector)){
  
  x <- as.character( eventVector[i] )
  
  prop_damage <- as.numeric(0) ## this shows damage in thousands!
  crop_damage <- as.numeric(0) ## this shows damage in thousands!
  
  for(j in 1:nrow(prop)){
    
      if(x == as.character(prop$event[j]) & as.character(prop$exp[j])=="B"){
        
        prop_damage <- prop_damage + prop$damage[j] * 1e+09
        
      }else if(x == as.character(prop$event[j]) ){
        
        prop_damage <- prop_damage + prop$damage[j] * 10^as.numeric( prop$exp[j] )
        
      }
     
  }
  
  for(j in 1:nrow(crop)){
    
    if(x == as.character(crop$event[j]) & as.character(crop$exp[j])=="B"){
      
        crop_damage <- crop_damage + crop$damage[j] * 1e+09
      
    }else if(x == as.character(crop$event[j]) ){
      
        crop_damage <- crop_damage + crop$damage[j] * 10^as.numeric( crop$exp[j] )
      
    }
    
  }
  
  total_damage = prop_damage + crop_damage
  
  additionalEventList <- list(x, prop_damage, crop_damage, total_damage )
  totalDmgDT <- rbind(totalDmgDT, additionalEventList)
  
}
```

<h4>5.3 Order events according to their damage</h4>

Step 5.3 orders the totalDmgDt and copies the results to the table economicImpact.

```{r}
economicImpact <- totalDmgDT[order(-total_dmg, -property_dmg, -crop_dmg)]
```

<h2>PART B. RESULTS</h2>

Part B now explains the results and shows the according figures.

<h3>6. Plot fatalities</h3>

Step 6 shows the number of fatalities associated with certain types of events. 
Due to the number of events (985), I limited it to the 10 events with associated with most fatalities. 
Interestingly, most fatalities are associated with tornados (5633) followed by excessive heats (1903).

```{r, results='asis' }
fatalitiesPlot <- healthImpact[1:10]

print( xtable(fatalitiesPlot), type="html"  )
```

```{r}
ggplot(data=fatalitiesPlot, aes(x=fatalitiesPlot$event, y=fatalitiesPlot$fatalities, fill="red") ) +
  theme_bw() + xlab("Event") + ylab("Fatalities") + geom_histogram(stat="identity") +
  theme(axis.text.x=element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_y_continuous(breaks=c(0, 1000, 2000, 3000, 4000, 5000, 6000), limits=c(0,6000))
```

<h3>7. Plot injuries</h3>

Step 7 shows the number of injuries associated with certain types of events. 
Due to the number of events (985), I limited it to the 10 events with associated with most injuries.
Again, most injuries are associated with tornados (91346). However, the followers are rather similar. 
That is, there are three events with injuries between 6500 and 7000.

```{r}
injuriesPlot <- healthImpact[1:10]

ggplot(data=injuriesPlot, aes(x=injuriesPlot$event, y=injuriesPlot$injuries, fill="blue") ) +
  theme_bw() + xlab("Event") + ylab("Injuries") + geom_histogram(stat="identity") +
  theme(axis.text.x=element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_y_continuous(breaks=c(0, 5000, 10000, 25000, 50000, 75000, 90000, 100000), limits=c(0,100000))
```

<h3>8. Plot economic damage</h3>

Finally, step 8 plots the economic damage. Again, tornadoes are leading. However, two types of floods are rather close and, 
if considered together, would have resulted in higher economic damage than tornados alone.

```{r, results='asis'}
damagePlot <- economicImpact[1:10]

print( xtable(damagePlot), type="html"  )

```

```{r}
ggplot(data=damagePlot, aes(x=damagePlot$event, y=damagePlot$total_dmg, fill="green") ) +
  theme_bw() + xlab("Event") + ylab("Economic Damage") + geom_histogram(stat="identity") +
  theme(axis.text.x=element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_y_continuous()
```
