A Study of an Individual's Daily Activity Pattern
========================================================

```{r, echo=FALSE}
options(warn=-1)
```

<h2>ABSTRACT</h2>
This work addresses the peer-assessed assignment number 1 for lecture "Reproducible Research" 
(Data Science Specialization, John Hopkins University).

<h2>Loading and preprocessing the data</h2>

First, I install and load the packages data.table, scales and ggplot2 for aggregation and visualization
of the data. Furthermore I read the data into the data.table activity make the dates to data type date.

```{r}
# Reproducible Research: Peer Assessment 1

## Load packages
##install.packages("data.table")
library(data.table)
##install.packages("scales")
library(scales)
##install.packages("ggplot2")
library(ggplot2)

## Loading and preprocessing the data
activity <- data.table( read.csv("C:/Users/Martin_Admin/Desktop/RepData_PA1/R_Project_PA1/activity.csv") )
activity$date <- as.Date(activity$date)
```

<h2>What is mean total number of steps taken per day?</h2>

I plot a histogram of the total number of steps taken per day. 

```{r}
## What is mean total number of steps taken per day?
stepsPerDay <- activity[, sum(steps), by="date"]
names(stepsPerDay)[2]<-paste("steps")

ggplot(data=stepsPerDay, aes(x=stepsPerDay$date, y=stepsPerDay$steps)) +
  theme_bw() + xlab("Date") + ylab("Steps") + geom_histogram(stat="identity")
```

Mean steps taken per day:
```{r}
mean(stepsPerDay$steps, na.rm=TRUE)
```

Median steps taken per day:
```{r}
median(stepsPerDay$steps, na.rm=TRUE)
```

<h2>What is the average daily activity pattern?</h2>

I define a datatable called intervalsSteps that will be filled with all intervals and their according total steps 
and number of occurences in the data. This will allow me to finally compute the average as total steps 
divided by days.

```{r}
## What is the average daily activity pattern?
intervalsSteps <- data.table(max1=numeric(0), max2=numeric(0), max3=numeric(0))
intervalsSteps <- rbind( intervalsSteps, list(-1, -1, -1) )

names(intervalsSteps)[1]<-paste("interval")
names(intervalsSteps)[2]<-paste("totalSteps")
names(intervalsSteps)[3]<-paste("intervalCount")
```

Now I fill the intervalsSteps datatable and in the end remove the helper line again.

```{r}

for(i in 1:nrow(activity) ){
  
  intervalInDtYet <- FALSE
  
  for(j in 1:nrow(intervalsSteps)){
    
    if( activity$interval[i] == intervalsSteps$interval[j] & !intervalInDtYet & !is.na(activity$steps[i]) ){
      
      intervalInDtYet <- TRUE ## interval is in DT yet
      
      intervalsSteps$totalSteps[j] <- intervalsSteps$totalSteps[j] + activity$steps[i]
      intervalsSteps$intervalCount[j] <- intervalsSteps$intervalCount[j] + 1
      
    }
    
  }
  
  if( !intervalInDtYet & !is.na(activity$steps[i]) ){ ## interval is not in DT yet
   
    additionalInterval = list( activity$interval[i], activity$steps[i], 1 )
    intervalsSteps <- rbind(intervalsSteps, additionalInterval)
    
  }
  
}

intervalsSteps2 <- intervalsSteps[intervalsSteps$interval>=0] ## remove helper line from the beginning
```

Now I compute all the averages and add them as an additional column to the 
intervalsSteps datatable. After that I draw a time series plot over the 
5-minute intervals (x-axis) and the average number of steps taken (averaged across 
all days; y-axis).

```{r}

intervalsAvgs <- data.table(max1=numeric(0))
names(intervalsAvgs)[1]<-paste("avgSteps")

for(i in 1:nrow(intervalsSteps2)){ 
  additionalAvg <- intervalsSteps2$totalSteps[i] / intervalsSteps2$intervalCount[i]
  intervalsAvgs <- rbind(intervalsAvgs, additionalAvg)
}

intervalsSteps2 <- cbind(intervalsSteps2, intervalsAvgs)

ggplot(data=intervalsSteps2, aes( x=intervalsSteps2$interval, y=intervalsSteps2$avgSteps) ) + 
  geom_line(stat="identity") + xlab("Daily 5-Minute Interval") + ylab("Avg. Steps")
```

I compute the interval with (on average) the most steps:

```{r}

maxAvgStepsInterval_i <- 1
for( i in 2:nrow(intervalsSteps2) ){
  
  if( intervalsSteps2$avgSteps[i] > intervalsSteps2$avgSteps[maxAvgStepsInterval_i] ){
    maxAvgStepsInterval_i <- i
  }
  
}
maxAvgStepsInterval <- intervalsSteps2$interval[maxAvgStepsInterval_i]
maxAvgStepsAmount <- intervalsSteps2$avgSteps[maxAvgStepsInterval_i]
```

The interval with on average most steps is:
```{r}
maxAvgStepsInterval
```

On average, the individual performed the following number of steps:
```{r}
mmaxAvgStepsAmount
```

<h2>Inputing missing values</h2>

I compute total number of missing values:
```{r}
countMissingValues <- 0

for(i in 1:nrow(activity)){
  if(is.na(activity$steps[i])){
    countMissingValues <- countMissingValues+1
  }
}
```

Subtask 1: Total number of missing values in the dataset:

```{r}
countMissingValues
```

Subtask 2 and 3: For filling in missing values, I compute the means for the respective 5-minute intervals.
Furthermore, I create a new dataset with the missing data filled in. This is stored under activityFilled.csv

```{r}
activityFilled <- activity

for(i in 1:nrow(activityFilled)){
  
  if(is.na(activityFilled$steps[i])){
    
    for(j in 1:nrow(intervalsSteps2)){
      
      if(activityFilled$interval[i]==intervalsSteps2$interval[j]){
        
        activityFilled$steps[i] <- intervalsSteps2$avgSteps[j]
        
      }
      
    }
    
  }
  
}
```

Storing the new dataset:

```{r}
write.csv(activityFilled, "activityFilled.csv", row.names=FALSE)
```

Subtask 4: I compute the mean and median of total number of steps taken per day. Also I draw a histogram of the total number of steps taken each day.

```{r}
stepsPerDayFilled <- activityFilled[, sum(steps), by="date"]
names(stepsPerDayFilled)[2]<-paste("steps")
```

The mean total number of steps taken per day is:

```{r}
mean(stepsPerDayFilled$steps, na.rm=TRUE)
```

The median total number of steps taken per day is:

```{r}
median(stepsPerDayFilled$steps, na.rm=TRUE)
```

Histogram indicating total number of steps taken per day:

```{r}
ggplot(data=stepsPerDayFilled, aes(x=stepsPerDayFilled$date, y=stepsPerDayFilled$steps)) +
  theme_bw() + xlab("Date") + ylab("Steps") + geom_histogram(stat="identity")

```

<h5>Interpretation of Mean and Median</h5>
The mean does not differ, because in the first part NAs have been ignored (as was demanded within the assignment).
However, the median increased (slightly), because empty rows now contain the average steps and do thus do less
"weaken" the difference between mean and median. (In fact, the median now is exactly the same as the mean!)
Finally, the histograms differ, because inputting the mean for missing intervals increases the number of steps
per day and thus increases the total number of steps per day. Regarding the histogram, a higher total number of
steps per day "enlarges" the bars in the histogram.

<h2>Are there differences in activity patterns between weekdays and weekends?</h2>

First, I create a new factor variable with the two levels weekday and weekend:

```{r}
activityFilledDaytype <- activityFilled
daytype <- data.table(max1=numeric(0))
names(daytype)[1]<-paste("daytype")

for(i in 1:nrow(activityFilledDaytype)){
  additionalDaytype <- weekdays(activityFilledDaytype$date[i])
  if(additionalDaytype=="Samstag" |
       additionalDaytype=="Saturday" |
       additionalDaytype=="Sonntag" |
       additionalDaytype=="Sunday"){
    additionalDaytype2<-"weekend"
  }else{
    additionalDaytype2<-"weekday"
  }
  daytype <- rbind(daytype, additionalDaytype2)
}

activityFilledDaytype <- cbind(activityFilledDaytype, daytype)

```

Now I create a table with the interval (filled with means) for all weekdays:

```{r}
intervalsStepsFilledWeekday <- data.table(max1=numeric(0), max2=numeric(0), max3=numeric(0))
intervalsStepsFilledWeekday <- rbind( intervalsStepsFilledWeekday, list(-1, -1, -1) )

names(intervalsStepsFilledWeekday)[1]<-paste("interval")
names(intervalsStepsFilledWeekday)[2]<-paste("totalSteps")
names(intervalsStepsFilledWeekday)[3]<-paste("intervalCount")

activityFilledWeekday <- activityFilledDaytype[activityFilledDaytype$daytype=="weekday"]

for(i in 1:nrow(activityFilledWeekday) ){
  
  intervalInDtYet <- FALSE
  
  for(j in 1:nrow(intervalsStepsFilledWeekday)){
    
    if( activityFilledWeekday$interval[i] == intervalsStepsFilledWeekday$interval[j] & 
          (!intervalInDtYet)) {
      
      intervalInDtYet <- TRUE ## interval is in DT yet
      
      intervalsStepsFilledWeekday$totalSteps[j] <- intervalsStepsFilledWeekday$totalSteps[j] + activityFilledWeekday$steps[i]
      intervalsStepsFilledWeekday$intervalCount[j] <- intervalsStepsFilledWeekday$intervalCount[j] + 1
      
    }
    
  }
  
  if(!intervalInDtYet){ ## interval is not in DT yet
    
    additionalInterval = list( activityFilledWeekday$interval[i], activityFilledWeekday$steps[i], 1 )
    intervalsStepsFilledWeekday <- rbind(intervalsStepsFilledWeekday, additionalInterval)
    
  }
  
}

intervalsStepsFilledWeekday2 <- intervalsStepsFilledWeekday[intervalsStepsFilledWeekday$interval>=0] ## remove helper line from the beginning

intervalsWeekdaysAvgs <- data.table(max1=numeric(0))
names(intervalsWeekdaysAvgs)[1]<-paste("avgSteps")

for(i in 1:nrow(intervalsStepsFilledWeekday2)){ 
  additionalAvg <- intervalsStepsFilledWeekday2$totalSteps[i] / intervalsStepsFilledWeekday2$intervalCount[i]
  intervalsWeekdaysAvgs <- rbind(intervalsWeekdaysAvgs, additionalAvg)
}

intervalsStepsFilledWeekday2 <- cbind(intervalsStepsFilledWeekday2, intervalsWeekdaysAvgs)

```

Now I create a table with the interval (filled with means) for all weekend-days:

```{r}
intervalsStepsFilledWeekend <- data.table(max1=numeric(0), max2=numeric(0), max3=numeric(0))
intervalsStepsFilledWeekend <- rbind( intervalsStepsFilledWeekend, list(-1, -1, -1) )

names(intervalsStepsFilledWeekend)[1]<-paste("interval")
names(intervalsStepsFilledWeekend)[2]<-paste("totalSteps")
names(intervalsStepsFilledWeekend)[3]<-paste("intervalCount")

activityFilledWeekend <- activityFilledDaytype[activityFilledDaytype$daytype=="weekend"]

for(i in 1:nrow(activityFilledWeekend) ){
  
  intervalInDtYet <- FALSE
  
  for(j in 1:nrow(intervalsStepsFilledWeekend)){
    
    if( activityFilledWeekend$interval[i] == intervalsStepsFilledWeekend$interval[j] & 
          (!intervalInDtYet)) {
      
      intervalInDtYet <- TRUE ## interval is in DT yet
      
      intervalsStepsFilledWeekend$totalSteps[j] <- intervalsStepsFilledWeekend$totalSteps[j] + activityFilledWeekend$steps[i]
      intervalsStepsFilledWeekend$intervalCount[j] <- intervalsStepsFilledWeekend$intervalCount[j] + 1
      
    }
    
  }
  
  if(!intervalInDtYet){ ## interval is not in DT yet
    
    additionalInterval = list( activityFilledWeekend$interval[i], activityFilledWeekend$steps[i], 1 )
    intervalsStepsFilledWeekend <- rbind(intervalsStepsFilledWeekend, additionalInterval)
    
  }
  
}

intervalsStepsFilledWeekend2 <- intervalsStepsFilledWeekend[intervalsStepsFilledWeekend$interval>=0] ## remove helper line from the beginning

intervalsWeekendAvgs <- data.table(max1=numeric(0))
names(intervalsWeekendAvgs)[1]<-paste("avgSteps")

for(i in 1:nrow(intervalsStepsFilledWeekend2)){ 
  additionalAvg <- intervalsStepsFilledWeekend2$totalSteps[i] / intervalsStepsFilledWeekend2$intervalCount[i]
  intervalsWeekendAvgs <- rbind(intervalsWeekendAvgs, additionalAvg)
}

intervalsStepsFilledWeekend2 <- cbind(intervalsStepsFilledWeekend2, intervalsWeekendAvgs)

```

Combine the datasets for weekdays and weekend-days:

```{r}
## Combine datasets and draw panel plot
intervalsStepsFilledAll <- rbind(intervalsStepsFilledWeekday2, intervalsStepsFilledWeekend2)

dayDT <- data.table(max1=character())
names(dayDT)[1]<-paste("daytype")

for(i in 1:nrow(intervalsStepsFilledWeekday2)){ 
  additionalDay <- "weekday"
  dayDT <- rbind(dayDT, additionalDay)
}
for(i in nrow(intervalsStepsFilledWeekday2)+1:nrow(intervalsStepsFilledWeekend2)){ 
  additionalDay <- "weekend"
  dayDT <- rbind(dayDT, additionalDay)
}

intervalsStepsFilledAll <- cbind(intervalsStepsFilledAll, dayDT)
```

Plot the intervals (filled with means) for all weekend-days in one panel plot:

```{r}
ggplot(data=intervalsStepsFilledAll, 
       aes( x=intervalsStepsFilledAll$interval, y=intervalsStepsFilledAll$avgSteps) ) + 
  geom_line(stat="identity") + xlab("Weekend 5-Minute Interval") + ylab("Avg. Steps") +
  facet_grid(daytype ~ .)
```

Clearly one can notice the differences between the averages on weekdays and weekends. While there appear to be some few peeks 
during weekdays (morning, lunch, afternoon), it is difficult to recognize a pattern for weekend-days.
